GWAS datasets for FDR benchmarking

### 2017-09-28 datasets/GWAS/GWAS.Rmd
Add diagnostic plots for covariates (Issue #13)

---
title: "GWAS datasets for FDR benchmarking"
author: "Keegan Korthauer"
date: "9/21/2017"
output: html_document
---

@@

# wrangling & plotting tools
library(data.table)
library(readxl)
library(dplyr)
library(ggplot2)

@@

@@

```{r, GWAS1-dat}
bmi <- bmi %>% 
        rename(pval = p_value,
               SE = stderr,
               effect_size = effect,
               ind_covariate = N) %>%
        mutate(zscore = effect_size/SE)

# Demo on a small subset of SNPs first (remove this later)
bmi <- bmi[sample(1:nrow(bmi), 100000),]
```

@@

@@

```{r, GWAS2-dat}
asthma <- asthma %>% 
        rename(pval = P_fix,
               SE = se_theta_fix,
               effect_size = theta_fix,
               ind_covariate = no_studies_meta) %>%
        mutate(zscore = effect_size/SE)

# Demo on a small subset of SNPs first (remove this later)
asthma <- asthma[sample(1:nrow(asthma), 100000),]
```

@@

@@

# remove SNPs with missing values for zscore (couldn't determine direction of 
# effect)
trig <- trig %>% 
  filter(!is.na(zscore))

# Demo on a small subset of SNPs first (remove this later)
trig <- trig[sample(1:nrow(trig), 100000),]
```

>>>>

---
title: "GWAS datasets for FDR benchmarking"
author: "Keegan Korthauer"
date: "9/28/2017"
output: html_document
---

@@

# wrangling & plotting tools
library(data.table)
library(readxl)
library(dplyr)
library(ggplot2)
library(cowplot)

@@

#### Covariate Diagnostics

Here we examine whether the available covariates appear to be suitable for the  
assumptions of the benchmark methods. The main assumption used by many methods 
is that the covariate is independent of the p-value conditional on the null
hypothesis. In other words, the covariate itself is not informative for whether
the test should be rejected or not. If this is satisfied, the covariate still
may or may not be useful. To assess whether the covariate is useful or not, we 
look to see if there is an association between covariate values and p-values. 
This is because under the alternative, we expect to see an enrichment of 
rejections for certain values of the covariate (if it is indeed informative).
In this section we examine two diagnostic plots for each of the covariates to 
see if these conditions hold in this dataset.

First, we will look at histograms of p-values stratified by covariate values. To
stratify by continuous covariate values, we'll use quantiles.

```{r, GWAS1-diag1, fig.width=9, fig.height=2.5}
# Demo on a small subset of SNPs first (remove this later)
bmi <- bmi[sample(1:nrow(bmi), 100000),]

# First for the sample size covariate (N)
strat_hist(bmi, pval="p_value", covariate="N", maxy=3)

# Next for the MAF covariate (Freq_Allele1_HapMapCEU)
strat_hist(bmi %>% filter(!is.na(Freq_Allele1_HapMapCEU)),
           pval="p_value", covariate="Freq_Allele1_HapMapCEU", maxy=3)

Next we'll look at a scatter plot of p-values by the ranks of covariate values.

```{r, GWAS1-diag2, fig.width=7, fig.height=4}
# First for the sample size covariate (N)
rank_scatter(bmi, pval="p_value", covariate="N")

# Next for the MAF covariate (Freq_Allele1_HapMapCEU)
rank_scatter(bmi, pval="p_value", covariate="Freq_Allele1_HapMapCEU")

# Take proximity to 0.5 as the actual covariate (according to previous plot)
bmi <- bmi %>%
        mutate(MAF.5=-abs(Freq_Allele1_HapMapCEU-0.5))
rank_scatter(bmi, pval="p_value", covariate="MAF.5") 
```

@@

```{r, GWAS1-dat}
bmi <- bmi %>% 
        rename(pval = p_value,
               SE = stderr,
               effect_size = effect,
               ind_covariate = N) %>%
        mutate(zscore = effect_size/SE)
```

@@

#### Covariate Diagnostics

First, we will look at histograms of p-values stratified by covariate values. To
stratify by continuous covariate values, we'll use quantiles.

```{r, GWAS2-diag1, fig.width=9, fig.height=2.5}
# Demo on a small subset of SNPs first (remove this later)
asthma <- asthma[sample(1:nrow(asthma), 100000),]

# First for the sample size covariate (no_studies_meta)
strat_hist(asthma, pval="P_fix", covariate="no_studies_meta", maxy=1.3)

# Next for the MAF covariate (freq_all_1_max)
strat_hist(asthma %>% filter(!is.na(freq_all_1_max)),
           pval="P_fix", covariate="freq_all_1_max", maxy=1.3)
```

Next we'll look at a scatter plot of p-values by the ranks of covariate values.

```{r, GWAS2-diag2, fig.width=7, fig.height=4}
# First for the sample size covariate (N)
rank_scatter(asthma, pval="P_fix", covariate="no_studies_meta")

# Next for the MAF covariate (freq_all_1_max)
rank_scatter(asthma, pval="P_fix", covariate="freq_all_1_max")

# Take distance to 0.5 as the actual covariate 
asthma <- asthma %>%
        mutate(MAF.5=abs(freq_all_1_max-0.5))
rank_scatter(asthma, pval="P_fix", covariate="MAF.5") 
```

@@

```{r, GWAS2-dat}
asthma <- asthma %>% 
        rename(pval = P_fix,
               SE = se_theta_fix,
               effect_size = theta_fix,
               ind_covariate = no_studies_meta) %>%
        mutate(zscore = effect_size/SE)
```

@@

#### Covariate Diagnostics

First, we will look at histograms of p-values stratified by covariate values. To
stratify by continuous covariate values, we'll use quantiles.

```{r, GWAS3-diag1, fig.width=9, fig.height=2.5}
# Demo on a small subset of SNPs first (remove this later)
trig <- trig[sample(1:nrow(trig), 100000),]

# First for the sample size covariate (no_studies_meta)
strat_hist(trig, pval="P-value", covariate="N", maxy=1.3)

# Next for the MAF covariate (Freq.A1.1000G.EUR)
strat_hist(trig %>% filter(!is.na(Freq.A1.1000G.EUR)),
           pval="P-value", covariate="Freq.A1.1000G.EUR", maxy=1.3)
```

Next we'll look at a scatter plot of p-values by the ranks of covariate values.

```{r, GWAS3-diag2, fig.width=7, fig.height=4}
# First for the sample size covariate (N)
rank_scatter(trig, pval="P-value", covariate="N")

# Next for the MAF covariate (Freq.A1.1000G.EUR)
rank_scatter(trig, pval="P-value", covariate="Freq.A1.1000G.EUR") 
```

@@

# remove SNPs with missing values for zscore (couldn't determine direction of 
# effect)
trig <- trig %>% 
  filter(!is.na(zscore))
```
